Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-06-13 docs/paper.Rmd
expand theoretical background w/ full income constraint and dynamic model (partial draft)

---
title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
short: "The Beckerian Quantity-Quality Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
vol: 1
issue: 1
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households. I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities. Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height. [Will add more after data analysis]
output: rticles::aea_article
bibliography: references.bib
---

@@

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2010fertility} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

This research contributes to the literature in several respects. It introduces new evidence on the nutritional consequences of fertility shocks in an urban developing-country context. It expands the measurement of child quality beyond educational outcomes to include biological indicators. It also adds to the limited body of work that uses plausibly exogenous variation in reproductive health policy to study downstream effects on human capital. More broadly, the findings underscore the potential for local family planning restrictions to generate unintended consequences for child well-being in settings characterized by poverty, food insecurity, and institutional fragility.

>>>>

---
title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households. I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities. Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height. [Will add more after data analysis]
output: 
  rticles::aea_article:
    keep_tex: true
    latex_engine: pdflatex
bibliography: references.bib
---

@@

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

This study contributes to the literature in several important ways. It provides new evidence on how increases in fertility—caused by policy restrictions on family planning—can affect child nutrition in poor, urban communities. Most past research has focused on education, but this study expands the idea of child quality to include biological measures such as stunting and wasting. It also adds to the small number of studies that use unexpected changes in reproductive health policy to examine long-term effects on children’s well-being. More broadly, the results show that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public services.

\section{Review of Related Literature}

\subsection{Theoretical Background}

Gary Becker’s early work reframed fertility as an economic decision and drew on microeconomic theory to how families choose whether and how many children to have. In his 1960 paper, Becker challenged the then-dominant view that fertility declines with income were simply the result of better access to contraception or changing social values. Instead, he proposed that childbearing decisions could be modeled using the tools of consumer choice theory. Within this framework, children are treated as goods that provide utility, with households allocating resources to balance their desire for more children (quantity) against the desire to invest in each child (quality).

Becker argued that children serve both consumption and production purposes. They bring direct satisfaction to parents and may also contribute economically, especially in agrarian or informal settings. In the Philippines, many children aged 5–17 work informally—through street vending or helping in family businesses—to support household income, with an estimated 1.09 million children working in 2023 \citep{manilabulletin2024}. This economic role aligns with Becker’s framework and may partly explain why some low-income households continue to prefer larger families. At the same time, fertility tends to decline as income rises and per-child investments increase. According to the 2017 National Demographic and Health Survey, women in the poorest wealth quintile had an average of 4.5 children, compared to just 2.0 in the richest \citep{psa2018}. This inverse relationship between income and fertility, observed globally, also appears in the Philippine context: poorer households often maintain larger families, while wealthier ones prioritize child quality.

\citet{beckerlewis1973} build on these conceptual foundations to introduce the formal concept of a quantity–quality (Q–Q) trade-off. They develop a utility-maximizing framework in which parents allocate resources between the number of children and the quality of investments in each child. The model formalizes how parents maximize utility from consumption $c$, number of children $n$, and child quality $q$:

$$
U = U(c, n, q)
$$

subject to the full income constraint: $$ I = c+n(p_n + q \times p_q) $$ where $p_n$ denotes the direct cost associated with bearing each additional child (e.g., delivery, basic needs), and $p_q$ represents the marginal cost of investing in quality per child, such as health care or education. The trade-off is reflected in the structure of the cost function, which reveals interaction effects: the marginal cost of improving child quality rises as the number of children increases, and similarly, the cost of an additional child rises with the level of quality investment per child. This complementarity is captured through the positive cross-derivatives:

$$
\frac{\partial MC_q}{\partial n} > 0 \quad \text{and} \quad \frac{\partial MC_n}{\partial q} > 0
$$

In short, the more children a family has, the more difficult it becomes to allocate sufficient resources—such as time, nutrition, and education—to each child.

While the static Q–Q model explains the immediate trade-offs families face between the number and quality of children, it remains limited in scope by assuming a single-period decision framework, which fails to account for intertemporal linkages—for instance, how early investments in child health or education may constrain or enhance later household choices—and overlooks how forward-looking behavior, such as anticipated returns to child quality in the form of future earnings or social mobility, may influence present-day household choices. To address this, \citet{becker1976child} developed a dynamic model that explicitly incorporates intergenerational mobility. In this model, parents account for how their fertility and investment choices shape the human capital outcomes of their children—for example, how a decision to space births more closely can constrain parental resources and reduce per-child investments in health and education. The resulting dynastic utility function captures both present and future-oriented preferences:

$$
U = u(c) + \beta \cdot n \cdot V(q)
$$

where $\beta$ denotes child altruism, or the extent to which parents value their children’s future well-being relative to their own consumption. A higher $\beta$ indicates stronger intergenerational concern, meaning parents are more willing to sacrifice current resources to improve their children’s prospects. For example, a low-income family in the Philippines might choose to send a child to private school despite financial constraints to increase the child’s future earnings or social mobility. Conversely, a lower $\beta$ implies a stronger emphasis on present needs and diminished willingness to invest in long-term child outcomes. This may occur in households facing severe income insecurity, where sending children to work may be prioritized over education. On the other hand, $V(q)$ represents the expected return on child quality. A high $V(q)$ implies that parents expect substantial future benefits from quality investments, such as access to better-paying jobs or improved life expectanc, making such investments more worthwhile than having additional children. A low $V(q)$, in contrast, suggests that even with significant investments, the perceived or actual gains are limited, often due to barriers like poor school quality, weak labor demand, or geographic constraints.

To formalize the trade-off similarly as in the static model, we again impose the full income constraint. In optimizing this problem, the first-order condition for child quality becomes:

$$
\beta \cdot V'(q) = p_q + n \cdot \frac{\partial p_q}{\partial q}
$$

This condition implies that the marginal benefit of child quality, scaled by altruism $\beta$, must equal the marginal cost of improving that quality. As family size $n$ increases, the cost of maintaining or increasing quality also rises. This reinforces the quantity–quality trade-off: families with more children may face sharper constraints in deepening investments per child. In contexts like the Philippines, where income constraints are binding and child quality has high private returns, this model predicts lower fertility and increased education spending among upwardly mobile households.

### 2025-06-13 docs/paper.tex
expand theoretical background w/ full income constraint and dynamic model (partial draft)

\date{\today}
\pubMonth{06}
\pubYear{2025}
\pubVolume{1}
\pubIssue{1}
\JEL{J13, I15, O15}
\Keywords{fertility shocks, child nutrition, quantity-quality
trade-off, urban poverty, Philippines}

@@

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2010fertility} to confirm that the contraceptive ban led to
an increase in fertility among affected women. I then estimate
reduced-form effects of policy exposure on nutritional outcomes for
children under five years of age. Finally, I examine heterogeneity in
effects across subsamples defined by maternal education, household
wealth, and access to prenatal care. These dimensions serve as proxies
for household resource availability and capacity to buffer the
nutritional consequences of fertility increases.

This research contributes to the literature in several respects. It
introduces new evidence on the nutritional consequences of fertility
shocks in an urban developing-country context. It expands the
measurement of child quality beyond educational outcomes to include
biological indicators. It also adds to the limited body of work that
uses plausibly exogenous variation in reproductive health policy to
study downstream effects on human capital. More broadly, the findings
underscore the potential for local family planning restrictions to
generate unintended consequences for child well-being in settings
characterized by poverty, food insecurity, and institutional fragility.

>>>>

\date{\today}
\pubMonth{06}
\pubYear{2025}
\pubVolume{}
\pubIssue{}
\JEL{J13, I15, O15}
\Keywords{fertility shocks, child nutrition, quantity-quality
trade-off, urban poverty, Philippines}

@@

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2019sex} to confirm that the contraceptive ban led to an
increase in fertility among affected women. I then estimate reduced-form
effects of policy exposure on nutritional outcomes for children under
five years of age. Finally, I examine heterogeneity in effects across
subsamples defined by maternal education, household wealth, and access
to prenatal care. These dimensions serve as proxies for household
resource availability and capacity to buffer the nutritional
consequences of fertility increases.

This study contributes to the literature in several important ways. It
provides new evidence on how increases in fertility---caused by policy
restrictions on family planning---can affect child nutrition in poor,
urban communities. Most past research has focused on education, but this
study expands the idea of child quality to include biological measures
such as stunting and wasting. It also adds to the small number of
studies that use unexpected changes in reproductive health policy to
examine long-term effects on children's well-being. More broadly, the
results show that local restrictions on family planning can
unintentionally harm children's health, especially in settings where
families already face poverty, food insecurity, and limited public
services.

\section{Review of Related Literature}

\subsection{Theoretical Background}

Gary Becker's early work reframed fertility as an economic decision and
drew on microeconomic theory to how families choose whether and how many
children to have. In his 1960 paper, Becker challenged the then-dominant
view that fertility declines with income were simply the result of
better access to contraception or changing social values. Instead, he
proposed that childbearing decisions could be modeled using the tools of
consumer choice theory. Within this framework, children are treated as
goods that provide utility, with households allocating resources to
balance their desire for more children (quantity) against the desire to
invest in each child (quality).

Becker argued that children serve both consumption and production
purposes. They bring direct satisfaction to parents and may also
contribute economically, especially in agrarian or informal settings. In
the Philippines, many children aged 5--17 work informally---through
street vending or helping in family businesses---to support household
income, with an estimated 1.09 million children working in 2023
\citep{manilabulletin2024}. This economic role aligns with Becker's
framework and may partly explain why some low-income households continue
to prefer larger families. At the same time, fertility tends to decline
as income rises and per-child investments increase. According to the
2017 National Demographic and Health Survey, women in the poorest wealth
quintile had an average of 4.5 children, compared to just 2.0 in the
richest \citep{psa2018}. This inverse relationship between income and
fertility, observed globally, also appears in the Philippine context:
poorer households often maintain larger families, while wealthier ones
prioritize child quality.

\citet{beckerlewis1973} build on these conceptual foundations to
introduce the formal concept of a quantity--quality (Q--Q) trade-off.
They develop a utility-maximizing framework in which parents allocate
resources between the number of children and the quality of investments
in each child. The model formalizes how parents maximize utility from
consumption \(c\), number of children \(n\), and child quality \(q\):

\[
U = U(c, n, q)
\]

subject to the full income constraint: \[ I = c+n(p_n + q \times p_q) \]
where \(p_n\) denotes the direct cost associated with bearing each
additional child (e.g., delivery, basic needs), and \(p_q\) represents
the marginal cost of investing in quality per child, such as health care
or education. The trade-off is reflected in the structure of the cost
function, which reveals interaction effects: the marginal cost of
improving child quality rises as the number of children increases, and
similarly, the cost of an additional child rises with the level of
quality investment per child. This complementarity is captured through
the positive cross-derivatives:

\[
\frac{\partial MC_q}{\partial n} > 0 \quad \text{and} \quad \frac{\partial MC_n}{\partial q} > 0
\]

In short, the more children a family has, the more difficult it becomes
to allocate sufficient resources---such as time, nutrition, and
education---to each child.

While the static Q--Q model explains the immediate trade-offs families
face between the number and quality of children, it remains limited in
scope by assuming a single-period decision framework, which fails to
account for intertemporal linkages---for instance, how early investments
in child health or education may constrain or enhance later household
choices---and overlooks how forward-looking behavior, such as
anticipated returns to child quality in the form of future earnings or
social mobility, may influence present-day household choices. To address
this, \citet{becker1976child} developed a dynamic model that explicitly
incorporates intergenerational mobility. In this model, parents account
for how their fertility and investment choices shape the human capital
outcomes of their children---for example, how a decision to space births
more closely can constrain parental resources and reduce per-child
investments in health and education. The resulting dynastic utility
function captures both present and future-oriented preferences:

\[
U = u(c) + \beta \cdot n \cdot V(q)
\]

where \(\beta\) denotes child altruism, or the extent to which parents
value their children's future well-being relative to their own
consumption. A higher \(\beta\) indicates stronger intergenerational
concern, meaning parents are more willing to sacrifice current resources
to improve their children's prospects. For example, a low-income family
in the Philippines might choose to send a child to private school
despite financial constraints to increase the child's future earnings or
social mobility. Conversely, a lower \(\beta\) implies a stronger
emphasis on present needs and diminished willingness to invest in
long-term child outcomes. This may occur in households facing severe
income insecurity, where sending children to work may be prioritized
over education. On the other hand, \(V(q)\) represents the expected
return on child quality. A high \(V(q)\) implies that parents expect
substantial future benefits from quality investments, such as access to
better-paying jobs or improved life expectanc, making such investments
more worthwhile than having additional children. A low \(V(q)\), in
contrast, suggests that even with significant investments, the perceived
or actual gains are limited, often due to barriers like poor school
quality, weak labor demand, or geographic constraints.

To formalize the trade-off similarly as in the static model, we again
impose the full income constraint. In optimizing this problem, the
first-order condition for child quality becomes:

\[
\beta \cdot V'(q) = p_q + n \cdot \frac{\partial p_q}{\partial q}
\]

This condition implies that the marginal benefit of child quality,
scaled by altruism \(\beta\), must equal the marginal cost of improving
that quality. As family size \(n\) increases, the cost of maintaining or
increasing quality also rises. This reinforces the quantity--quality
trade-off: families with more children may face sharper constraints in
deepening investments per child. In contexts like the Philippines, where
income constraints are binding and child quality has high private
returns, this model predicts lower fertility and increased education
spending among upwardly mobile households.
